Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Abstract

\begin{abstract}

@@

\ADComment{Open to attack: why not just buy another GPU?}

@@

\ADComment{TODO: justify our choice: uniform variables and efficiency.}

@@

\ADComment{Perhaps this would fit better under ``programming guidelines'' -- not sure.}

\ADComment{Make the point here that if an irregular algorithm is
  written as a cooperative kernel then it will execute just as safely
  as it does today on current hardware if: (1) the $\transmit$
  annotation and $\offerkill$/$\offerfork$ calls are defined away to
  nothing, (2) $\resizingglobalbarrier$ is redefined to
  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
  that assumes the occupancy-bound execution model is provided.  If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}

>>>>

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  Current irregular algorithms only work on today's
GPUs by identifying hardware compute units and scheduling workgroups
exclusively to these compute units.  As well as being non-portable,
this is unacceptable because the GPU cannot be used in the meantime
for high-priority tasks with soft real-time constraints,
such as graphics rendering.

We propose \emph{cooperative kernels}, an extension to the traditional
GPU programming model geared towards writing blocking algorithms.
Cooperative kernels enable fair scheduling of workgroups executing a
blocking algorithm, as well as co-scheduling of other tasks on the
GPU, via a small set of language extensions that create a contract
between the programmer and the scheduler.  We describe our cooperative
kernels programming model and its semantics in detail, and our
implementation of a portable prototype of cooperative kernels on top
of OpenCL 2.0.  We evaluate the technique by porting a set of
irregular work stealing and graph algorithms to our programming model
and evaluate the performance and responsiveness of the approach across
four GPUs from two vendors.  Because our implementation exploits no
vendor-specific hardware, driver or compiler support, our encouraging
experimental results provide a lower-bound on the efficiency with which
cooperative kernels can be implemented in practice.

@@

@@

We opted for transmitting only selected varibles based on the
observation that many of a thread's private variables are dead at the
point of issuing $\offerfork$ or $\resizingglobalbarrier$, thus it
would be wasteful to transmit them.  A live variable analysis could
instead be employed to over-approximate the variables that might be
accessed by newly arriving workgroups, so that these are automatically
transmitted.

In all cases, we found that a variable that needed to be transmitted
had the property of being uniform across the workgroup.  That is,
despite each thread having its own copy of the variable, each thread
is in agreement on the variable's value.  As an example, the
$\keyword{level}$, $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$
variables used in \myfig\ref{fig:cgraphsearch} are all stored in thread-private
memory, but all threads in a workgroup agree on the values of these
variables at each $\resizingglobalbarrier$ call.  As a result,
transmitting the thread 0's valuation of the annotated variables is
equivalent to (and more efficient than) transmitting values on a
thread-by-thread basis.  We have not yet encountered a real-world
example where our current semantics would not suffice.

@@

A goal of coopeative kernels is to allow traditional, non-blocking GPU
tasks, such as graphics shaders (in programming models such as OpenGL,
Vulkan and DirectX) and regular data-parallel computational kernels,
to be co-scheduled with blocking irregular algorithms.  It is
important to note that our new model has no effect on the manner in
which these traditional GPU workloads are programmed.

Furthermore, if our cooperative language extensions are
ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
$\transmit$ annotations are ignored, and $\globalbarrier$ and
$\resizingglobalbarrier$ calls are redirected to an existing global
barrier implementation (so that no resizing occurs in the latter
case)---then cooperative kernel will behave just like its
non-cooperative counterpart would behave.  As a result, a developer
for whom today's occupancy-bound execution model suffices can upgrade
their kernel to a cooperative form and retain backwards-compatibility
with today's GPU platforms by pre-processing away our language
extensions.
